---
title: "Long Term S&P 500 Models"
author: "Micah Williams"
date: "3/18/2021"
output: html_document
---

```{r setup, include=FALSE}
# set options
knitr::opts_chunk$set(echo = F,
                      fig.width = 7.5,
                      fig.height = 5,
                      fig.align = 'center')

# import required packages
library(tidyverse)
library(RColorBrewer)
library(lubridate)

theme_set(theme_minimal())

# import S&P500 data from Quandl package
sp <- Quandl::Quandl("MULTPL/SP500_INFLADJ_MONTH", api_key="ouXsKHk7DUeaYiFanm9b") %>%
  janitor::clean_names() %>% 
  arrange(date) %>%
  
  # filter out rows with inconsistent date
  mutate(day = day(date)) %>% 
  filter(day == 1) %>%
  select(-day) %>%
  
  # add row numbers for easier id
  mutate(id = row_number())

# function to find monthly value change
findChange <- function(){
  vals <- c()
  prices <- sp[,'value']
  for (i in c(1:nrow(sp))){
    change <- prices[i+1] / prices[i]
    vals <- append(vals, change)
  }
  # return vector of values
  tibble('factor' = vals)
}

# add monthly change column
sp <- bind_cols(sp, findChange()) %>% 
  mutate('pct_change' = (factor - 1) * 100) %>%
  
  # reorder columns
  .[,c(3,1,2,4,5)]

# import recession dates
recessions <- read_csv('data/nber_recessions.csv') %>% 
  filter(end > min(sp$date) | is.na(end)) %>%
  mutate(end = if_else(is.na(end), as.Date('2021-06-01'), end))
```

```{r structure, fig.width= 7.5}
head(sp[,-3])
```


This dataset contains monthly inflation-adjusted (in January 2021 dollars) average closing prices of the S&P 500 index. There are **1,803** rows in the dataset, and each row contains a date, the first day of a month, and a value, the average closing price of the index during the month ending on the given day. The dates range from Jan. 1, 1871 to March 1, 2021. The average closing prices range from \$70.77 to \$3901.82.

Here's what the data looks like with the y-axis on a log scale, with recessions marked with shaded areas:

```{r prices}
ggplot() +
  
  geom_rect(aes(xmin = start, xmax = end,
                ymin = 70.77, ymax = 3901.82),
            fill = 'red',
            alpha = 0.15,
            data = recessions) +
  
  geom_line(aes(date, value),
            color = 'navy',
            size = 1.2,
            data = sp) +
  
  labs(title = 'Average S&P500 Closing Price by Month, from Jan. 1, 1871 to March 1, 2021',
       subtitle = 'Shaded areas indicate recessions',
       caption = 'Price data from Quandl, Recession dates from NBER.\nValues are inflation adjusted to Jan. 2021') +
  
  scale_y_log10(name = 'Average Closing Price (log scale)') +
  scale_x_date(breaks = as.Date(paste0(seq(1875,2025,25),'-01-01')),
               date_labels = '%Y',
               name = '')
```

## Time in the Market

```{r longterm}

# function to calculate with intermediate values
buySP <- function(start, end = '2021-03-01', amount = 10000){
  
  tbl <- sp %>% 
    filter(between(date, as.Date(start), as.Date(end)))
  
  tbl %>%
    mutate(investment = map_dbl(value, ~ amount * . / tbl[1,'value']))
  
}

# only calculates start and end
buySP_quick <- function(start, end = '2021-03-01', amount = 10000){
  
  tbl <- sp %>% 
    filter(date %in% c(as.Date(start), as.Date(end)))
  
  tbl %>%
    mutate(investment = map_dbl(value, ~ amount * . / tbl[1,'value']))
  
}

# extract final value from table
endval <- function(tbl){
  tbl %>% slice(nrow(tbl)) %>% pull(investment)
}

# all in on Jan 1 1871
endval(buySP_quick('1871-01-01'))

# wait for the crash
endval(buySP_quick(sp[sp$value == min(sp$value),'date']))
```

Now, let's pretend you have \$10,000 to invest. If you invested it all on Jan. 1 1871, you'd have around \$419,000 today. A handsome return! It's a shame you're too dead to enjoy it. 

If you had waited for the market to crash, as it did a few years later, and bought in when the S&P was at its lowest (down around 25% from Jan. 1871), you'd have more than \$551,000. You'd be the richest dead guy in the cemetary!

Of course, you probably want to enjoy your wealth while you're still alive, and 150 years is a long time to wait. If you only had 40 years to invest, what month would've been the best to start?

```{r returns}
buySP_40 <- function(start, amount = 10000){
  
  # find id of starting month
  start_id <- sp[sp$date == as.Date(start),'id']
  
  tbl <- sp %>% 
    
    # filter to start month and 40 years later
    filter(id %in% c(start_id, start_id + 480))
  
  # calculate return
  tbl %>% mutate(investment = map_dbl(value, ~ amount * . / tbl[1,'value']))
  
}

get_return <- function(date){
  endval(buySP_40(date))
}

# filter for possible start months for 40 year period
forty_years <- sp %>% 
  slice(c(1:(nrow(sp) - 480))) %>%
  mutate(return = map_dbl(date, ~ get_return(.)))

# plot returns
ggplot() +
  
  geom_rect(aes(xmin = start, xmax = end,
                ymin = 0, ymax = 100000),
            fill = 'gray',
            alpha = 0.5,
            data = recessions[recessions$start <= as.Date('1981-03-01'),]) +
  
  geom_area(aes(date, return),
            color = 'forestgreen',
            fill = 'forestgreen',
            size = 1.2,
            alpha = 0.5,
            data = forty_years) +
  
  geom_ribbon(aes(date, 
                  ymin = return,
                  ymax = 10000),
            color = 'red',
            fill = 'red',
            size = 1.2,
            alpha = 0.5,
            data = forty_years[forty_years$return <= 10000,]) +
  
  geom_hline(aes(yintercept = 10000),
             color = '#333333',
             size = 1.2) +
  
  labs(title = 'Value of $10,000 Invested for 40 Years in S&P500 Index, by start date',
       subtitle = 'Shaded areas indicate recessions',
       y = 'Value 40 years later',
       x = 'Start date of investment',
       caption = 'Price data from Quandl, Recession dates from NBER.\nValues are inflation adjusted to Jan. 2021') +
  
  scale_y_continuous(breaks = seq(0,10^5,25000),
                     labels = paste0('$', seq(0,100,25), 'k')) + 
  scale_x_date(breaks = as.Date(paste0(seq(1870,1980,20),'-01-01')),
               date_labels = '%Y')
```

If you picked a random month in this range to invest $10,000 in an S&P500 index you'd turn a profit more than 80% of the time. And you'll notice a trend in the areas where the overall return was negative: these were months where their 40-year investment interval included the Depression of 1920-21 (following World War I) or the Great Depression (1929-33) or both. Referring back to the price chart shows that you would've been closing up your investment when prices were near all-time lows after buying at relatively high prices.

Most of the local peaks occur in months where recessions end and prices are bottoming out.

Towards the right side of the graph, things just seem to keep going up and up, with the exception of a brief period in the mid-1970's, where the investment interval would end in the middle of the 2008 Financial Crisis. However, if you held strong while the market bottomed out, you'd come out clean on the other side. Interestingly, while the American economy is currently in a recession, the stock market has continued its march upwards and has recently hit multiple all-time highs.

## Timing the Market

Now, you might look at the graph above and think, "Well, why don't I just wait for the market to crash, and I'll buy when things are cheaper!" You certainly wouldn't be alone in that logic. Let's say you keep the 40 year investment interval, but instead of investing everything at the beginning, you wait and try to buy when you think the prices are at their lowest.

With the simple math above, it wouldn't make much of a difference- the only penalty for waiting to invest is rising prices. But, in the real world, how long you keep an investment matters much more because some stocks pay dividends, which can be reinvested. In this way, investors can use compound interest to grow the value of their investment exponentially. In reality, the value of dividends changes just like the stock, but to simplify things, let's assume an annual 2% dividend, and that we'll immediately reinvest the dividend. For example, if you invest \$10,000, and the index increases 10%, at the end of the first year you'll have \$11,220: \$10,000 principal, \$1,000 (10%) appreciation, and a 2% (\$220) dividend on your new principal of $11,000.

```{r compounding}

# calculate returns with annual dividend
compound <- function(start, end = '2021-03-01', amount = 10000, dividend = 0.02){
  
  # extract relevant data
  tbl <- sp %>% 
    filter(between(date, as.Date(start), as.Date(end)))
  
  # vector for returns
  returns <- tibble('investment' = rep(0, length.out = nrow(tbl)))
  
  # calculate returns by row
  for(i in c(1:nrow(tbl))){
    
    # find change in price, add dividend every 12 months
    returns[i, 'investment'] <- amount * tbl[i, 'value'] / tbl[1, 'value'] * (1 + dividend) ^ (floor(i/12))
  }
  
  # return table with returns column
  bind_cols(tbl, returns)
}

# calculate returns with annual dividend (without intermediate values)
compound_quick <- function(start, end = '2021-03-01', amount = 10000, dividend = 0.02){
  
  # extract relevant data
  tbl <- sp %>% 
    filter(date %in% c(as.Date(start), as.Date(end)))
  
  # vector for returns
  returns <- tibble('investment' = rep(0, length.out = nrow(tbl)))
  
  # calculate returns by row
  for(i in c(1:nrow(tbl))){
    
    # find change in price, add dividend every 12 months
    returns[i, 'investment'] <- amount * tbl[i, 'value'] / tbl[1, 'value'] * (1 + dividend) ^ (floor(i/12))
  }
  
  # return table with returns column
  bind_cols(tbl, returns)
}

# function to calculate returns after 40 years without intermediate values
compound40 <- function(start, amount = 10000, dividend = 0.02){
  
  # find id of starting month
  start_id <- sp[sp$date == as.Date(start),'id']
  
  # extract data for relevant dates
  tbl <- sp %>% 
    
    # filter to start month and 40 years later
    filter(id %in% c(start_id, start_id + 480))
  
  # vector for returns
  returns <- tibble('investment' = rep(0, length.out = nrow(tbl)))
  
  # calculate return by row
  for(i in c(1:nrow(tbl))){
    
    # find change in price, add dividend every 12 months
    returns[i, 'investment'] <- amount * tbl[i, 'value'] / tbl[1, 'value'] * (1 + dividend) ^ 40
  }
  
  # return table with returns column
  bind_cols(tbl, returns)
}

# pull end dates of recessions
recession_ends <- sp %>% 
  filter(date %in% recessions$end) %>%
  select(id, date)

buy_the_dip <- function(start){
  
  # find id of input
  input_id <- sp[sp$date == as.Date(start), 'id']
  
  find 
  
  # find end date of next recession and investment period (40 years)
  start_date <- recession_ends[min(which(recession_ends$id >= input_id)), 'date']
  end_date <- sp[sp$id == input_id + 480, 'date']
  
  # calculate return from start date to 40 years after input date
  results_tbl <- compound(start = start_date, end = end_date)
  
  # extract final value
  results_tbl
}

# calculate compound returns for 40 years
forty_years_compound <- forty_years %>%
  slice_sample(prop = 0.05) %>%
  mutate(compound = map_dbl(date, ~endval(compound40(.))),
         dip = map_dbl(date, ~endval(buy_the_dip(.)))) %>%
  
  # pivot columns to tidy format
  pivot_longer(return:dip, names_to = 'method', values_to = 'return') %>%
  
  # change descriptors
  mutate(method = case_when(method == 'return' ~ 'No Compounding',
                            method == 'compound' ~ 'Compounding',
                            method == 'dip' ~ 'Buy the Dip'))

# plot returns
ggplot() +
  
  geom_rect(aes(xmin = start, xmax = end,
                ymin = 0, ymax = 150000),
            fill = 'gray',
            alpha = 0.5,
            data = recessions[recessions$start <= as.Date('1981-03-01'),]) +
  
  geom_line(aes(date, return,
                color = method),
            # fill = 'forestgreen',
            size = 1.2,
            alpha = 0.5,
            data = forty_years_compound) +
  
  scale_color_manual(values = brewer.pal(3, 'Dark2')[c(1,3,2)]) +
  
  geom_hline(aes(yintercept = 10000),
             color = '#333333',
             size = 1.2) +
  
  labs(title = 'Value of $10,000 Invested for 40 Years in S&P500 Index, by start date',
       subtitle = 'Compounding calculated with 2% annual dividend. Shaded areas indicate recessions',
       y = 'Value 40 years later',
       x = 'Start date of investment',
       caption = '\'Buy The Dip\': set an end date of start + 40 yrs, but waiting until end of next recession to invest.\nPrice data from Quandl, Recession dates from NBER. Values are inflation adjusted to Jan. 2021',
       color = 'Method') +
  
  scale_y_continuous(breaks = seq(0,max(forty_years_compound$return),25000),
                     labels = paste0('$', seq(0, max(forty_years_compound$return) / 10^4, 25), 'k')) +
  scale_x_date(breaks = as.Date(paste0(seq(1870,1980,20),'-01-01')),
               date_labels = '%Y') +
  
  theme(legend.position = c(0.2, 0.7),
        legend.background = element_rect(fill = '#ffffff8d',
                                         color = '#adadad'))
```

